\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-14 coqimpl.tex
Save.

\textbf{Family compiled into module.} This is the main idea of compiling family components into Coq's primitives: we compile family into coq module, family type into module type and context of a given term into parameters of the module. To achieve incremental type checking, we notice the nature of late binding: for each defining field $ .. \cL\sigma \vdash t : T $, we can compile $t$ using ``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall \cL \sigma. T$. However, instead of using Coq's universal quantifier, we use module and module parameters to achieve this wrapping -- each family type in the context of the judgement will be one module parameter with the compiled module type, and thus each term of the field will compile into a parametrized module (functor). Doing so we can also get rapid feedback from type checking when the users are defining each field interactively with Coq. Overriding is achieved in the same manner -- we just need to make sure the overriding term has the same type as the original term.

% Concrete details of compilation of a family term data structure

% Explain the whole procedure using this pseudo-code example
Now that we know a family data structure is actually a list of compiled modules, we need to aggregate them into a complete compiled module. Thankfully, Coq's \textit{Include} command makes things easy and reasonably fast. One key feature of Coq's \textit{Include} is that, it can include not-completely instantiated functor (i.e. module with non-instantiated parameters) and it will try to find appropriate fields in the current surrounding module to instantiate those parameters. This feature can solve the fact that each defining field requires a ``self'' parameter.

% Compilation into Coq can give us computational interpretation for free.

% explaining inheritance
To simplify the implementation, we can consider each family as an inheritance and the standalone family are extending empty family. Thus we directly construct the data structure representing inheritance during interactive proving. An inheritance will fix the type for the input family and output family, and generally has three kinds of inheritance in OO language -- extend, override and inherit. Extending corresponds to adding new fields, overriding corresponds to modify existent fields and inheriting corresponds to directly taking fields from the parent. It is possible to consider inheritance as solely a data rather than a functor, and this actually provides us the ability to \text{mix-in} inheritance data.

\textbf{Reasoning Requires Non-Overridable.}
Unlike general-purposed programming, Coq should be able to  \textit{reason} about each program. However, sometimes reasoning requires a specific program and overriding on them is unacceptable. For example, we may define a field $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field and intends to proving its commutativity in a following field $"comm"$. But to prove so, we need to know more than just a type interface of $"add"$ but also its \textbf{concrete definition}. Otherwise, in the future, once $"add"$ is overridden into a subtraction function then $"comm"$ will have problem to be  inherited. To achieve such exposure of definition, in meta-theory we only need to expose the information as propositional equality as part of the signature, and in our Coq plugin, we can achieve that even easier by exposing the whole definition directly into module type. Doing so we can have a better judgemental equality and we can avoid overriding because subtraction doesn't have the same definition as $add$. We require the users to decorate field as `Final' so that the plugin will proceed with this special treatment.

>>>>

\textbf{Family compiled into module.} This is the main idea of compiling family components into Coq's primitives: we compile family into coq module, family type into module type and context of a given term into parameters of the module. To achieve incremental type checking, we notice the nature of late binding: for each defining field $ .. \cL\sigma \vdash t : T $, we can compile $t$ using ``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall \cL \sigma. T$. However, instead of using Coq's universal quantifier, we use module and module parameters to achieve this wrapping -- each family type in the context of the judgement will be one module parameter with the compiled module type, and thus each term of the field will compile into a parametrized module (functor). Doing so we can also get rapid feedback from type checking when the users are defining each field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b 
  : self_A.a = self_A.a
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptySig.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

Module A. (* Final Aggregation *)
Include a_4. Include b_7. End A.
\end{minted}
  \end{minipage}
  \caption{Example Code and Plugin Translation}\label{fig:plugin-example1}
\end{figure}

% Concrete details of compilation of a family term data structure
Taking Figure \ref*{fig:plugin-example1} as an example: when the user interactively input the line "Field a : nat := ..", our plugin will translate this statement into the "Module a_9", "Module Type a_10, a_11" three components. {The "EmptySig" is just empty module type.} These modules and module types will be type-checked by Coq on the fly. The computed "Module a_10" is then part of the context of "Field b" (as the module type of "self__A") for "b" to refer to. The internal representation of family "A" is simply the list of these ``compiled'' modules and module types. We can simply achieve incremental checking and overriding -- for example, we want to override "Field a", we simply replace that compiled module with the new module inside the list.

Now that we know a family data structure is actually a list of compiled modules, we may want to aggregate them into one complete compiled module. We achieve that by repeatedly using Coq's "Include" command as the example illustrated. Note that, in our example, we include two functors without arguments -- this is due to one of the feature of "Include": non-instantiated parameters will be automatically instantiated by the appropriate fields in the surrounding defining module. The resulting module is just like mundane Coq's module and thus enjoy all the computational mechanism from Coq for free.

% explaining inheritance
Besides standalone family, we also need to implement family inheritance. To simplify the implementation, we can consider each family as an inheritance and the standalone family are extending empty family. Thus we directly construct the data structure representing inheritance during interactive proving.

An inheritance data will be annotated with the type for the input family and output family, and generally there are three kinds of inheritance -- extend, override and inherit. Extending corresponds to adding new fields, overriding corresponds to modify existent fields and inheriting corresponds to directly taking fields from the parent. Thus it is possible to consider inheritance as solely a data rather than a functor, and this actually provides us the ability to \text{mix-in} inheritance data.

\textbf{Reasoning Requires Non-Overridable.}
Unlike general-purposed programming, Coq should be able to  \textit{reason} about each program. However, sometimes reasoning requires a specific program and overriding on them is unacceptable. For example, we may define a field $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field and intends to proving its commutativity in a following field $"comm"$. But to prove so, we need to know more than just a type interface of $"add"$ but also its \textbf{concrete definition}. Otherwise, in the future, once $"add"$ is overridden into a subtraction function then $"comm"$ will have problem to be  inherited. A more concrete example is in Figure \ref*{fig:plugin-example1}, consider we have "Field b : self_A.a = 1 := .." instead, then our compilation and Coq's type check will both fail: in the context of "b" (module type "a_6") we only know "a" is of type nat but nothing more.

To achieve such exposure of definition, in meta-theory we only need to expose the information as propositional equality as part of the signature, and in our Coq plugin, we can achieve that even easier by exposing the whole definition directly into module type. Doing so we can have a better judgemental equality and we can avoid overriding because subtraction doesn't have the same definition as $add$. We require the users to decorate field as `Final' so that the plugin will proceed with this special treatment.

### 2022-06-14 standardmodel.tex
Save.

  \denotesS{\goodType{\Gamma}{\cL \sigma}{i}} \ (\gamma)&= \{\star\} \text{ when } \denotesS{\sigma} \in \denotesS{\goodSig{\Gamma}{\_}{0}} \\

@@

  \denotesS{\goodTerm{\Gamma}{\mu\cdot}{\cL \nu\cdot} } \ (\gamma) &= \star \\

@@

@@

The verification of validity of this model is largely omitted here and we refer the reader to the appendix. To verify our model preserve the judgemental equality, usually unpacking the definition is enough -- for example, to verify $\beta$-rule for linkage projection, we only need to see that $\denotesS{\goodTerm{\Gamma}{p_1\mu~(\mu^+~o~t)}{\cL\sigma}} = \denotesS{\goodTerm{\Gamma}{o}{\cL\sigma}}$. One exception is for $\cC$ and $\goodSig{\_}{\_}{n}$, the substitution law around these two are also proved together by induction because they were mutual-recursively defined together. The compatibility lemmas can be verified directly -- for example $\frac{\goodSigS{\Gamma}{\sigma}{n+1}\quad\goodTermS{\Gamma}{o}{\cL\sigma}}{\goodTermS{\Gamma}{p_1\mu~o}{\cL(p_1\nu~\sigma)}}$ is proved by simply unfolding the definition.

>>>>

  \denotesS{\goodType{\Gamma}{\cL \sigma}{i}} (\gamma)&= \{\star\} \text{ when } \denotesS{\sigma} \in \denotesS{\goodSig{\Gamma}{\_}{0}} \\

@@

  \denotesS{\goodTerm{\Gamma}{\mu\cdot}{\cL \nu\cdot} } (\gamma) &= \star \\

@@

  \denotesS{\goodTerm{\Gamma}{p_1\mu~t}{\cL(p_1\nu~\sigma)}}(\gamma) &=  "fst"~(t\gamma) \\ 

@@

The verification of validity of this model is largely omitted here and we refer the reader to the appendix. To verify our model preserve the judgemental equality, usually unpacking the definition is enough -- for example, to verify $\beta$-rule for linkage projection, we only need to see that $\denotesS{\goodTerm{\Gamma}{p_1\mu~(\mu^+~o~t)}{\cL\sigma}} = \denotesS{\goodTerm{\Gamma}{o}{\cL\sigma}}$ by function extensionality. One exception is for $\cC$ and $\goodSig{\_}{\_}{n}$, the substitution law around these two are also proved together by induction because they were mutual-recursively defined together. The compatibility lemmas can be verified directly -- for example $\frac{\goodSigS{\Gamma}{\sigma}{n+1}\quad\goodTermS{\Gamma}{o}{\cL\sigma}}{\goodTermS{\Gamma}{p_1\mu~o}{\cL(p_1\nu~\sigma)}}$ is proved by simply unfolding the definition.
